---
title: "Online Shopping Analysis"
author: "S20820"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries


```{r warning=FALSE, include=FALSE}

library(dplyr)
library(janitor)
library(factoextra)
library(corrplot)
library(ggplot2)
library(tidyverse)
library(CCA)
library(heplots)
library(MASS)
library(caret)

set.seed(820)

```


## Loading dataset and preparation

## Loading dataset

```{r}
dataset <- read.csv("./online_shoppers_intention.csv")

dataset <- clean_names(dataset)

dataset <- dataset %>% mutate(across(c(operating_systems, browser, region, traffic_type, visitor_type, weekend, revenue, month), as.factor))
dataset <- dataset %>% mutate(across(c(administrative, administrative_duration, informational, informational_duration, 
                                       product_related, product_related_duration, bounce_rates, exit_rates, page_values, special_day), as.numeric))

categorical_variables <- colnames(dataset %>% dplyr::select(where(is.factor)))
numerical_variables <- colnames(dataset %>% dplyr::select(where(is.numeric)))

numerical_data <- dataset %>% dplyr::select(all_of(numerical_variables))
categorical_data <- dataset %>% dplyr::select(all_of(categorical_variables))

str(dataset)                              

summary(dataset)

```

### Missing Value Handling

```{r}
cat("Number Of rows with missing values:", sum(dataset %>% is.na()))

```
- The dataset does not contain any missing values.


## Exploratory Data Analysis and Data Visualization


### Distribution of Numerical variables

```{r}
numerical_data %>% pivot_longer(cols = numerical_variables) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ name, scales = "free", ncol=4) +
  theme_minimal()


```




### Distribution of Categorical variables

```{r}


p <- categorical_data %>%
  pivot_longer(cols = -revenue, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = revenue)) +
  geom_bar(position = "fill") +
  facet_wrap(~ variable, scales = "free_x", ncol=2) +
  theme_minimal() +
  labs(title = "Proportion of Purchases across Categorical Variables",
       x = "Category Level",
       y = "Proportion",
       fill = "Made Purchase (Revenue)") +
  scale_fill_manual(values = c("tomato", "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=8))
p

# ggsave("categorical_plots.png", plot = p, width = 10, height = 10)

```

### Correlation

```{r}

cor_matrix <- cor(numerical_data)

corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Numerical Variables", mar=c(0,0,1,0))

```

### Outlier Handling

```{r}

numeric_vars_long <- scale(numerical_data) %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

ggplot(numeric_vars_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 8) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(title = "Boxplots of Standardized Numerical Variables",
       y = "Standardized Value", x = "Variable")




```




```{r}

is_outlier_iqr <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  
  return(x < lower_bound | x > upper_bound)
}

outlier_matrix <- sapply(numerical_data, is_outlier_iqr)

rows_with_any_outlier <- rowSums(outlier_matrix) > 0

total_rows <- nrow(dataset)
outlier_rows <- sum(rows_with_any_outlier)
percentage_removed <- (outlier_rows / total_rows) * 100

cat("Total rows originally:", total_rows, "\n")
cat("Rows flagged for removal:", outlier_rows, "\n")
cat(sprintf("Percentage of data that would be removed: %.2f%%\n", percentage_removed))


```
Using Inter quartile range method to remove outliers will remove 57.36% of the data which
is a significant amount of data.
Therefore the outliers will not be removed.


### Distribution of the target variable

```{r}

ggplot(dataset, aes(x = revenue, fill = revenue)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Distribution of Revenue (Target Variable)",
       x = "Revenue (Made a Purchase)", y = "Count") +
  scale_fill_manual(values = c("chartreuse4", "brown3"))

```





## Principal Component Analysis

```{r warning=FALSE}

numerical_data <- dataset %>% dplyr::select(all_of(numerical_variables))

pca_result <- prcomp(numerical_data, scale = TRUE)


print("Mean:")
pca_result$center
print("Variance:")
pca_result$scale



```

```{r}

summary(pca_result)
# According to Kaizer criteria choose 4 PCs
variances <- pca_result$sdev^2
print(variances)

```




```{r}
pca_result$rotation



```




```{r}
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50))
```


```{r}

fviz_pca_var(pca_result,
 col.var = "contrib", 
 gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
 repel = TRUE 
 )


```

```{r}
var <- get_pca_var(pca_result)
corrplot(var$cos2, is.corr=FALSE, col=COL2('RdBu'))

```

```{r}
loadings <- var$coord
print(loadings)


```




```{r}


scaled_data <- as.data.frame(scale(numerical_data))


set1 <- scaled_data %>% dplyr::select(administrative, informational, product_related)

set2 <- scaled_data %>% dplyr::select(bounce_rates, exit_rates, page_values)

cca_result <- cc(set1, set2)

cat("Canonical Correlations:\n")
print(cca_result$cor)

cat("Coefficients for Set 1 (Actions):\n")
print(cca_result$xcoef)
cat("Coefficients for Set 2 (Analytics):\n")
print(cca_result$ycoef)

plt.cc(cca_result, var.label = TRUE)
```

```{r}

library(CCP)
rho <- cca_result$cor
n <- nrow(dataset)
p <- ncol(set1)
q <- ncol(set2)
p.asym(rho, n, p, q, tstat = "Wilks") #


```



```{r}

u_scores <- cca_result$scores$xscores
v_scores <- cca_result$scores$yscores


cv_data <- data.frame(
  CV1_Set1 = u_scores[,1], 
  CV1_Set2 = v_scores[,1],
  Revenue = as.factor(dataset$revenue) 
)


ggplot(cv_data, aes(x = CV1_Set1, y = CV1_Set2)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  xlim(-8, 0.9) +
  theme_minimal() +
  labs(
    title = "Scatterplot of First Canonical Variate Pair",
    subtitle = paste("Canonical Correlation =", round(cca_result$cor[1], 4)),
    x = "Canonical Variate 1 (User Actions)",
    y = "Canonical Variate 1 (Site Analytics)"
  )

```






```{r}

cat("\n--- Discriminant Analysis Assumption Checks ---\n")

lda_data <- scaled_data
lda_data$revenue <- as.factor(dataset$revenue)

box_m_result <- boxM(lda_data[, 1:10], lda_data$revenue)
print(box_m_result)


lda_model <- lda(revenue ~ ., data = lda_data)
print(lda_model)

lda_predictions <- predict(lda_model)


confusion_matrix <- table(Predicted = lda_predictions$class, Actual = lda_data$revenue)
cat("\nConfusion Matrix:\n")
print(confusion_matrix)

accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat(sprintf("\nLDA Classification Accuracy: %.2f%%\n", accuracy * 100))

lda_plot_data <- data.frame(LD1 = lda_predictions$x[,1], Revenue = lda_data$revenue)

ggplot(lda_plot_data, aes(x = LD1, fill = Revenue)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "LDA: Density Plot of Discriminant Scores",
       x = "Linear Discriminant 1 (LD1)",
       y = "Density") +
  scale_fill_manual(values = c("tomato", "steelblue"))





```

```{r}
set.seed(4033)
train_index <- createDataPartition(lda_data$revenue, p = 0.8, list = FALSE)
training_data <- lda_data[train_index, ]
test_data <- lda_data[-train_index, ]

cat("\n--- Step 1: Training Data Collected ---\n")
cat("Training set size:", nrow(training_data), "\n")
cat("Test (Unknown) set size:", nrow(test_data), "\n")


prior_probs <- prop.table(table(training_data$revenue))
cat("\n--- Step 2: Estimated Prior Probabilities ---\n")
print(prior_probs)


cat("\n--- Step 3: Variance-Covariance Check ---\n")
cat("Note: Box's M-test previously yielded p < 0.05. \n")
cat("Heterogeneous matrices confirmed. QDA is the appropriate method.\n")

cat("\n--- Steps 4 & 5: QDA Model Estimation ---\n")

qda_model <- qda(revenue ~ ., data = training_data, prior = as.numeric(prior_probs))


print(qda_model)

cat("\n--- Step 6: Cross-Validation (Leave-One-Out) ---\n")

qda_cv <- qda(revenue ~ ., data = training_data, CV = TRUE)

confusion_table_cv <- table(Truth = training_data$revenue, Classified_As = qda_cv$class)
print(confusion_table_cv)

misclass_rate <- 1 - sum(diag(confusion_table_cv)) / sum(confusion_table_cv)
cat(sprintf("Estimated overall misclassification probability: %.2f%%\n", misclass_rate * 100))

cat("\n--- Step 7: Classifying Unknown Observations ---\n")


qda_predictions <- predict(qda_model, newdata = test_data)


cat("Predictions for the first 5 'unknown' specimens:\n")
head(data.frame(
  Actual_Revenue = test_data$revenue,
  Predicted_Class = qda_predictions$class,
  Posterior_Prob_FALSE = qda_predictions$posterior[,1],
  Posterior_Prob_TRUE = qda_predictions$posterior[,2]
), 5)

final_confusion <- table(Truth = test_data$revenue, Classified_As = qda_predictions$class)
final_accuracy <- sum(diag(final_confusion)) / sum(final_confusion)
cat(sprintf("\nFinal QDA Model Accuracy on Unknown Data: %.2f%%\n", final_accuracy * 100))
```

```{r}


train_index <- createDataPartition(lda_data$revenue, p = 0.8, list = FALSE)
training_data <- lda_data[train_index, ]
test_data <- lda_data[-train_index, ]



# Step 2: Estimated Prior Probabilities [cite: 1546-1555]
# Calculated from training data to reflect real-world imbalance
prior_probs <- prop.table(table(training_data$revenue))

# Step 3: Variance-Covariance Check (Performed via Box's M Test)
# Assumption of homogeneity failed (p < 0.05), justifying QDA [cite: 1563-1564]

# Steps 4 & 5: QDA Model Estimation [cite: 1569-1576]
# Fitting the model using estimated priors
qda_model <- qda(revenue ~ ., data = training_data, prior = as.numeric(prior_probs))

# Step 6: Cross-Validation (Leave-One-Out) [cite: 624-628]
qda_cv <- qda(revenue ~ ., data = training_data, CV = TRUE)

confusion_table_cv <- table(Truth = training_data$revenue, Classified_As = qda_cv$class)

# 1. Calculate Group-Specific Misclassification Probabilities p(i|j) [cite: 616-618]
# Probability that a unit from population j is classified into population i
misclass_probs <- prop.table(confusion_table_cv, margin = 1) 

cat("\n--- Group-Specific Misclassification Probabilities ---\n")
print(misclass_probs)

# 2. Calculate Overall Misclassification Probability [cite: 628]
overall_error <- 1 - (sum(diag(confusion_table_cv)) / sum(confusion_table_cv))
cat(sprintf("\nOverall Misclassification Probability: %.2f%%\n", overall_error * 100))

# Step 7: Classifying Unknown Observations [cite: 1585-1588]
# Applying model to the test set
qda_predictions <- predict(qda_model, newdata = test_data)

# --- RELEVANT PLOT: Density of Posterior Probabilities ---
# This plot visualizes the model's confidence in predicting a purchase (Revenue = TRUE)
plot_data <- data.frame(
  Prob_TRUE = qda_predictions$posterior[, "TRUE"],
  Actual_Revenue = test_data$revenue
)

ggplot(plot_data, aes(x = Prob_TRUE, fill = Actual_Revenue)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "QDA: Distribution of Posterior Probabilities for Purchase Intent",
    subtitle = "Visualizing class separation and model certainty",
    x = "Posterior Probability of Revenue = TRUE",
    y = "Density",
    fill = "Actual Revenue"
  ) +
  scale_fill_manual(values = c("tomato", "steelblue"))
```
```{r}
d_data <- as.data.frame(pca_result$x[, 1:4])
d_data$revenue <- dataset$revenue

# Step 2: Data Partitioning (80% Train, 20% Test)
set.seed(4033) 
train_idx <- createDataPartition(d_data$revenue, p = 0.8, list = FALSE)
train <- d_data[train_idx, ]
test <- d_data[-train_idx, ]

# Step 3: Assumption Check on PCs (Box's M Test) [cite: 460-461]
# Re-testing homogeneity because PCA changes the variance structure.
box_m_pc <- boxM(d_data[, 1:4], d_data$revenue)
cat("\n--- Box's M Test for PC Predictors ---\n")
print(box_m_pc)

# Step 4 & 5: QDA Model Estimation using PC Scores
# Note: PCA components are orthogonal, which stabilizes the estimation.
qda_model <- qda(revenue ~ ., data = train)

# Step 6: Cross-Validation (LOOCV)
qda_cv <- qda(revenue ~ ., data = train, CV = TRUE)
conf_matrix_cv <- table(Truth = train$revenue, Classified = qda_cv$class)
cat("\n--- PC-based QDA Cross-Validation Table ---\n")
print(conf_matrix_cv)

# Step 7: Final Accuracy on Unknown PC Data
qda_preds <- predict(qda_model, newdata = test)
final_acc_pc <- sum(diag(table(test$revenue, qda_preds$class))) / nrow(test)
cat(sprintf("\nFinal PCDA Accuracy: %.2f%%\n", final_acc_pc * 100))

plot_data <- data.frame(
  Prob_TRUE = qda_preds$posterior[, "TRUE"],
  Actual_Revenue = test$revenue
)

ggplot(plot_data, aes(x = Prob_TRUE, fill = Actual_Revenue)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "QDA: Distribution of Posterior Probabilities for Purchase Intent",
    subtitle = "Visualizing class separation and model certainty",
    x = "Posterior Probability of Revenue = TRUE",
    y = "Density",
    fill = "Actual Revenue"
  ) +
  scale_fill_manual(values = c("tomato", "steelblue"))
```

```{r}

# ==============================================================================
# DISCRIMINANT ANALYSIS: ONLINE SHOPPER INTENTION
# Method: Principal Component Discriminant Analysis (PCDA) using QDA
# ==============================================================================

# --- Step 1: Data Preparation & PC Extraction --- [cite: 446-447]
# Extract the first 4 Principal Components (capturing ~72% variance)
d_data <- as.data.frame(pca_result$x[, 1:4])
d_data$revenue <- dataset$revenue

# Split into Training (80%) and Test (20%) sets [cite: 617-619]
# Training set is for model building (Steps 2-6); Test set for Step 7.
set.seed(4033) 
train_idx <- createDataPartition(d_data$revenue, p = 0.8, list = FALSE)
training_set <- d_data[train_idx, ]
test_set <- d_data[-train_idx, ]

# --- Step 2: Prior Probabilities --- [cite: 449-451]
# Calculate estimated prior probabilities from the training data [cite: 457-458]
prior_probs <- prop.table(table(training_set$revenue))
cat("\n--- Step 2: Estimated Prior Probabilities ---\n")
print(prior_probs)

# --- Step 3: Assumption Check - Homogeneity of Covariance --- [cite: 459-461]
# Performing Box's M Test (Multivariate extension of Bartlett's Test)
# Null Hypothesis (H0): Covariance matrices are equal across groups [cite: 462-463]
box_m_pc <- boxM(training_set[, 1:4], training_set$revenue)
cat("\n--- Step 3: Box's M Test for Homogeneity ---\n")
print(box_m_pc) 
# Note: If p < 0.05, we reject H0 and MUST use QDA [cite: 466-468].

# --- Step 4: Assumption Check - Multivariate Normality --- [cite: 477]
# Visual diagnostic using a Chi-Square Q-Q Plot of Mahalanobis Distances
mu <- colMeans(training_set[, 1:4])
sigma <- var(training_set[, 1:4])
d2 <- mahalanobis(training_set[, 1:4], center = mu, cov = sigma)

qqplot(qchisq(ppoints(nrow(training_set)), df = 4), d2,
       main = "Step 4: Chi-Square Q-Q Plot for PC Normality",
       xlab = "Theoretical Chi-Square Quantiles",
       ylab = "Sample Mahalanobis Distances",
       pch = 20, col = "steelblue")
abline(0, 1, col = "tomato", lwd = 2)

# --- Step 5: Compute Discriminant Functions (QDA) --- [cite: 478, 664-666]
# Heterogeneous matrices (Step 3) justify Quadratic Score Functions [cite: 694]
qda_model <- qda(revenue ~ ., data = training_set, prior = as.numeric(prior_probs))

# --- Step 6: Misclassification Probabilities (Cross-Validation) --- [cite: 480-482]
# Using Leave-One-Out Cross-Validation (LOOCV) for unbiased estimates [cite: 624-628]
qda_cv <- qda(revenue ~ ., data = training_set, CV = TRUE)
conf_matrix_cv <- table(Truth = training_set$revenue, Classified_As = qda_cv$class)

# Calculate group-specific and overall misclassification rates [cite: 615-618]
misclass_probs <- prop.table(conf_matrix_cv, margin = 1)
overall_error <- 1 - (sum(diag(conf_matrix_cv)) / sum(conf_matrix_cv))

cat("\n--- Step 6: Cross-Validation Error Metrics ---\n")
print(misclass_probs)
cat(sprintf("Overall Misclassification Probability: %.2f%%\n", overall_error * 100))

# --- Step 7: Classify Unknown Observations --- [cite: 488]
# Applying the finalized model to the 20% "Unknown" test set [cite: 700-704]
final_preds <- predict(qda_model, newdata = test_set)

# Final Accuracy Check
final_conf <- table(Truth = test_set$revenue, Predicted = final_preds$class)
final_acc <- sum(diag(final_conf)) / sum(final_conf)
cat(sprintf("\n--- Step 7: Final Test Accuracy: %.2f%% ---\n", final_acc * 100))

# Visualization: Distribution of Posterior Probabilities
plot_df <- data.frame(Prob_TRUE = final_preds$posterior[, "TRUE"], 
                      Actual = test_set$revenue)

ggplot(plot_df, aes(x = Prob_TRUE, fill = Actual)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "QDA: Classification Confidence (Posterior Probabilities)",
       x = "Probability of Purchase (Revenue = TRUE)", y = "Density") +
  scale_fill_manual(values = c("tomato", "steelblue"))

```

```{r}

d_data <- as.data.frame(pca_result$x[, 1:4])
d_data$revenue <- dataset$revenue


train_idx <- createDataPartition(d_data$revenue, p = 0.8, list = FALSE)
train <- d_data[train_idx, ]
test <- d_data[-train_idx, ]


box_m_pc <- boxM(train[, 1:4], train$revenue)
print(box_m_pc)


mu <- colMeans(train[, 1:4])
sigma <- var(train[, 1:4])
d2 <- mahalanobis(train[, 1:4], center = mu, cov = sigma)

qqplot(qchisq(ppoints(nrow(train)), df = 4), d2,
       main = "Chi-Square Q-Q Plot for PCs",
       xlab = "Theoretical Chi-Square Quantiles",
       ylab = "Sample Mahalanobis Distances",
       pch = 20, col = "steelblue")
abline(0, 1, col = "tomato", lwd = 2)


qda_model <- qda(revenue ~ ., data = train)

qda_cv <- qda(revenue ~ ., data = train, CV = TRUE)
conf_matrix_cv <- table(Truth = train$revenue, Classified = qda_cv$class)

print(prop.table(conf_matrix_cv, margin = 1)) 

overall_error_pc <- 1 - (sum(diag(conf_matrix_cv)) / sum(conf_matrix_cv))
cat(sprintf("Overall CV Misclassification Probability: %.2f%%\n", overall_error_pc * 100))

qda_preds <- predict(qda_model, newdata = test)
final_acc_pc <- sum(diag(table(test$revenue, qda_preds$class))) / nrow(test)
cat(sprintf("\nFinal PCDA Accuracy: %.2f%%\n", final_acc_pc * 100))

plot_data <- data.frame(
  Prob_TRUE = qda_preds$posterior[, "TRUE"],
  Actual_Revenue = test$revenue
)

ggplot(plot_data, aes(x = Prob_TRUE, fill = Actual_Revenue)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Posterior Probabilities for Purchase Intent",
    x = "Posterior Probability of Revenue = TRUE",
    y = "Density",
    fill = "Actual Revenue"
  ) +
  scale_fill_manual(values = c("tomato", "steelblue"))

```

